---
title: "p8105_hw2_mjd2283"
author: "Max"
date: "2026-10-09"
output: github_document
---
#Problem 1


```{r}

library(tidyverse)

transit_df = 
  read_csv(
    "./NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) |>
  mutate(entry = case_match(entry, "Yes" ~ TRUE, "NO" ~ FALSE))
transit_df

#used col_types since when I first ran it, it doesn't have route 8 through 11 as chr so made this change

#used case_match to change entry to TRUE or FALSE

stations = transit_df |>
  distinct(line, station_name) |>
  nrow()

```

`transit_df` contains 1868 rows showing an entrance or exit and 19 columns. At each one, the dataset tells us the `line`, `station_name`, `station latitude`,`station_longitude`, route numbers among other information. So far, I've converted all my route#s to be characters since 8 through 11 were not. In addition, I selected certain variables to retain as mentioned in the problem. I also used case_match to change the variable entry from yes and no to true and false. 




#Problem 2

```{r}
library(tidyverse)

mr_trash_df = 
  readxl::read_excel("./202610 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N742") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    trash_wheel = "Mr Trash")

professor_trash_df =
  readxl::read_excel("./202610 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M139") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash")

gwynnda_df = 
  readxl::read_excel("./202610 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda the Good Wheel of the W", range = "A2:L394") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(trash_wheel = "Gwynnda")

combined_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df) |> 
  relocate(trash_wheel)

combined_df

#went from A2 headers to N742 last row of possible data excluding the totals row
#also added the sheet option
#clean names makes it snake case

#only including dumpster rows (drop_na), mutating so that the number of sports balls is rounded
#labeling each trash wheel as different names

#years can't combine in between sheets so need to fix
#also similar issue with plastic bottles

#clarifying names
```
```{r}

total_weight_professor = 
  combined_df |>
  filter(trash_wheel == "Professor Trash") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

print(total_weight_professor)

total_cigarette_gwynnda =
  combined_df |> 
  filter(trash_wheel == "Gwynnda", month == "June", year == "2022") |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
  
print(total_cigarette_gwynnda)

```

The combined dataset has `r nrow(combined_df)` observations and `r ncol(combined_df)` categories from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Key variables in this combined dataset include `weight_tons`, `volume_cubic_yards`, `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `plastic_bags`, `wrappers`, `sports_balls`, and `homes_powered`. Also information on month, date, and year is given. The total weight collected by Professor Trash Wheel was 249.96 tons, and the total number of cig butts collected by Gwynnda in June of 2022 was 18120. 

#Problem 3

```{r}

library(tidyverse)

baseline_original=
  read_csv("./data_mci/MCI_baseline.csv", 
    skip = 1, na = c (".", "NA", "")) |>
    janitor::clean_names() 

baseline_df =
  baseline_original |>
  mutate(
    sex = case_match(
        sex, 
        1 ~ "male", 
        0 ~ "female"),
    sex = as.factor(sex),
    apoe4 = case_match(
      apoe4, 
        1 ~ "carrier", 
        0 ~ "non-carrier"),
    apoe4 = as.factor(apoe4)) |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

#using case match to ensure not numeric 
  
#labeled it as characters and then used as.factor to turn it to a factor

#used filter() to remove participants who do not meet stated inclusion criteria

recruited = nrow(baseline_original)
included = nrow(baseline_df)
developed_mci = 
  baseline_df |>
  filter(!is.na(age_at_onset)) |>
  pull(id) |>
  length()
mean_age = mean(baseline_df$current_age, na.rm = TRUE)

baseline_df
  
#For this question, kept people with non missing values for age at onset. They developed MCI. Used AI to help with this approach

women_apoe4 = baseline_df |>
filter(sex == "female") |>
  mutate (gene = apoe4 == "carrier") |>
  pull(gene) |>
  mean()

#finding proportion by classifying APOE4 as either True or False so 1 or 0 and then finding the mean. Used AI to help with code reasoning here

amyloid_original=
  read_csv("./data_mci/MCI_amyloid.csv",
    skip = 1, na = c (".", "NA", "")) |>
  janitor::clean_names() |>
  rename(id = study_id) 

amyloid_df =
  amyloid_original |>
  pivot_longer(baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio") |>
   mutate(
    visit = replace(visit, visit == "baseline", "time_0"),
    visit = factor(visit)) 

amyloid_df

#pivot_longer to change so visits all under one column and values under another

n_baseline_only =
  baseline_df |>
  filter(!(id %in% amyloid_df$id)) |>
  pull(id) |>
  length()

n_baseline_only

n_amyloid_only =
  amyloid_original |>
  filter(!(id %in% baseline_df$id)) |>
  pull(id) |>
  length()

n_amyloid_only

#used %in% so see if the id value was in the other dataset and ! keeps only those that are not

mci_df = inner_join(baseline_df, amyloid_df, by = "id")

write_csv(mci_df, "./data_mci/mci_combined.csv")
  
mci_df

```
The `baseline_df` is 479 id participants long and includes variables such as current_age, sex, education, APOE4, and age_at_onset. In comparison, the `amyloid_df` is  2435 rows long representing id participants at baseline (`time_0`),`time_2`, `time_4`, `time_6,` and `time_8`. 

The `mci_df` contains id values across visits including time_0 which is baseline and then up to time_8, while keeping the overall structure of `baseline_df` with id being the first column. `mci_df` is 2355 rows representing 471 IDs with 5 visits each. 

8 participants are labeled only in the `n_baseline_only`. 16 participants are labeled only in the `n_amyloid_only`.






